---
title: "BEAST2 UCED date shuffling analyses"
subtitle: "Redefining the treponemal history through pre-Columbian genomes from Brazil"
author: "Louis du Plessis"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  github_document:
      toc: true
      toc_depth: 3
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: false
    fig_crop: false
params: 
  truthpath   : "../results/beast2_dateshuffling/xml/output/"
  shufflepath : "../results/beast2_dateshuffling/xml/output/"
  outputpath  : "../results/beast2_dateshuffling/"
  filename    : "all_01_02_23_alignment_norec.gtrgi.bsp10.uced.narrow"
  burnin      : 0.1
  reps        : 50
  device      : "png"


---


```{r rsetup, include=FALSE}
  
    library(cowplot)
    library(coda)
    library(beastio)
    source("../scripts/HPDBeanPlot.R")

    figpath <- paste0(params$outputpath, "figures/")
    cachepath <- paste0(params$outputpath, "cache/")

    dir.create(figpath, recursive = TRUE, showWarnings = FALSE)
    
    knitr::opts_chunk$set(tidy=FALSE, cache=FALSE, results='hide', cache.path = cachepath, 
                          dev=params$device, fig.path=figpath, fig.width=7.5, fig.height=4,
                          message=FALSE, error=FALSE, warning=FALSE, echo=FALSE)

    
```


```{r functions} 


#' Assume rates is a coda mcmc object 
plotRateComparison <- function(truerate, rates, names=NULL, xlabshift=-0.1,
                               maxwidth=0.5, bw='sj', mostrecent=2020, monophyletic=NULL, palfn=ggsci::pal_nejm, ylab="s/s/y", ylim=NULL, ...) {

      par(mar=c(3,5,1,1))

      rates   <- mcmc(cbind(truerate, rates))
      rateHPD <- getHPDMedian(rates)
      
      n        <- ncol(rates)
      pal      <- palfn()(8)
      paltrans <- palfn(alpha=0.5)(8)
      
      if (is.null(ylim)) {
        ylim <- c(0, max(rateHPD)*1.05)
      }
      yrange <- diff(ylim)

      plot(1, type='n', bty='n', xlim=c(0,n+1), ylim=ylim, axes=FALSE, xlab='', yaxs='i', xaxs='i', ylab=ylab, ...)
      axis(2, las=1, cex.axis=0.6)
      axis(1, at=0:n, labels = NA, lwd.ticks = NA)
      #abline(v = 1:n, lwd=0.25, lty=3)
      #abline(h = axTicks(2), lwd=0.25, lty=3)
      text(1:n, y = ylim[1]+xlabshift*yrange, names, srt=90, xpd=TRUE, cex=0.6)

      abline(h = rateHPD[1,c(1,3)], col=pal[1], lty=2)
      abline(h = rateHPD[1,2],      col=pal[1], lty=2)      
      HPDBeanPlot(rates, fill=paltrans[2], border=pal[2], medcol=pal[1], bw=bw, maxwidth=maxwidth, add=TRUE, axes=FALSE, lwd=c(0.5,1,0.5))

      #abline(h = mean(rates[,1]), lty=2, col=pal[4])
      #abline(h = max(rateHPD[2:nrow(rateHPD),]), lty=2, col=pal[4])
      
      #legend("top", horiz=TRUE, inset=c(0,-0.15), bty='n', fill=c(paltrans[c(3,2)]), border=c(pal[c(3,2)]), legend=legend, xpd=TRUE, cex=0.6)
}


extractFromMCMC.list <- function(trace.list, varname) {
    result <- mcmc(sapply(1:nchain(trace.list), function(i) trace.list[[i]][,varname]))
    varnames(result) <- chanames(trace.list)
    return(result)
}


```


\clearpage

# Methods

## Dataset

Sampling dates of ancient samples were fixed to the mean of the radiocarbon date range. 


## Model

Used UCED model with dates fixed to the mean radiocarbon age.

## Analyses
All analyses were performed in BEAST v2.6.7. MCMC chains were run for 500 million steps and parameters sampled every 10,000 steps. 10% of samples were discarded as burn-in.


\clearpage

# Results

## Convergence

```{r shuffleddates-uced, cache=TRUE, eval=TRUE}

  # Load data
  shufflereps <- list.files(path=params$shufflepath, pattern = "*.log")
  trace.shuffled <- readLog(paste0(params$shufflepath, shufflereps), burnin = params$burnin)
  
  # Check convergence
  minESSs <- sapply(1:length(shufflereps), function(i) min(checkESS(trace.shuffled[[i]], cutoff=Inf)))

```


```{r shuffleddates-convergence}

  plot(1, type='n', bty='n', xlim=c(0,length(minESSs)+1), ylim=c(0, max(minESSs)*1.05), axes=FALSE, xlab='', yaxs='i', xaxs='i', ylab="Lowest ESS value")
  axis(2, las=1, cex.axis=0.6)
  axis(1, at=0:length(minESSs), labels = NA, lwd.ticks = NA)
  text(1:length(minESSs), y = -0.1*max(minESSs), rep(NA, length(minESSs)), srt=90, xpd=TRUE, cex=0.6)
  
  points(1:length(minESSs), sort(minESSs, decreasing = TRUE), pch=16, col='navy')
  abline(h = 200, col='yellow', lty=2)
  abline(h = 150, col='red', lty=2)
  abline(v = params$reps, col='red', lty=2)
      
```
   
\clearpage
   
## Rate comparison
      
```{r shuffleddates-ucedrates, fig.cap="Date shuffling analyses performed under a **relaxed clock** models. The plot shows the posterior distributions for the mean clock rate, truncated at the upper and lower limits of the 95% HPD interval. Horizontal red lines indicate the medians of the posterior distributions. The red dashed lines indicate the median and upper and lower limits of the 95% HPD interval of the clock rate inferred under the true sampling dates.", eval=TRUE}
  
  trace.truth    <- readLog(paste0(params$truthpath, params$filename,".truth.121.log"), burnin = params$burnin)

  # Plot rate comparison
  rates.shuffled <- extractFromMCMC.list(trace.shuffled, "rateStat.mean")

  # Random selection with ESS > 150
  # selected <-  sample(which(minESSs > 150), params$reps)
  
  # Top ESSs
  selected <- order(minESSs, decreasing = TRUE)[1:params$reps]

  plotRateComparison(trace.truth[, "rateStat.mean"],  rates.shuffled[, selected], names=c("Truth", rep(NA,params$reps)))
  
```


\clearpage

# Session info

```{r sessionInfo, results='markup'}
    sessionInfo()
```
